### 2016-03-10 paper/icfp16-submission/semantics.tex
rewrite lemma k-stuck

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial input type upto trace $\trace$ will
cause $f$ to get stuck in \emph{at most} $k$
steps, where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  For all $v$,
  if \steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
     \vincompat{v}{\ptype{\trace}{f}},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
     in at most $k$ steps, where $k$ is the length of $\trace$.
\end{lem}
\begin{proof}
By induction on $k$, the length of $\trace$.
%
Suppose {\vincompat{v}{\ptype{\trace}{f}}}.
%
We show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
in at most $k$ steps.
%
The base case, $k = 0$ is trivial, $\trace$ is empty
and so $\ptype{\trace}{f}$ is a hole that is compatible
with \emph{every} value $v$.
%
In the inductive case, let
$\trace' = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e'}{\vsu'}{\tsu'}$
be the prefix of $\trace$ of length $k-1$.
%
Furthermore, let $s_{\trace} = \ptype{\trace}{f}$ and $s_{\trace'} = \ptype{\trace'}{f}$.
%
Let us split cases on whether $v$ is compatible with $s_{\trace'}$.
%
\begin{description}
\item [Case \vincompat{v}{s_{\trace'}}:]
  The inductive hypothesis applies.

\item [Case $\vcompat{v}{s_{\trace'}}$ but $\vincompat{v}{s_{\trace}}$:]
  Since $\vcompat{v}{s_{\trace'}}$ but $\vincompat{v}{s_{\trace}}$ we know
  that $s_{\trace'} \neq s_{\trace}$.
  By Lemma~\ref{lem:force-inst} we know that we must have
  invoked \forcesym\ at step $k$.
  %
  The Preservation Lemma~\ref{lem:refine-partial} implies that
  $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
  specifically narrowed $s_{\trace'}$ to a type incompatible with $v$.
  %
  A case analysis of the evaluation rules shows that such an
  invocation of \forcesym\ at step $k$
  cannot succeed, \ie\ yields \stuck.
   \RJ{some intuition needed --- seems like key step.}
\end{description}
\end{proof}

\begin{proof}[\textbf{Proof of Theorem~\ref{thm:soundness}}]
%
Suppose $\trace$ witnesses that $f$ gets stuck,
and let $t = \ptype{\trace}{f}$.
We show that \emph{all} types $s$ have stuck-inducing
values by splitting cases on whether the type is
compatible with $t$. %the partial type upto $\trace$.
%
\begin{description}
\item [Case \tcompat{s}{t}:]
  Let $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{\stuck}{\vsu}{\tsu}$.
  %
  The value $v = \resolve{\ehole}{\vsu}$ demonstrates that
  $\eapp{f}{v}$ gets stuck.
\item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$
  such that \hastype{v}{t} demonstrates that $\eapp{f}{v}$ gets stuck.
  % \ES{do we need to say anythign else?}
\end{description}
\end{proof}

>>>>

\begin{lem}
\label{lem:vsu-ext}
If  $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$
and $\trace'$ is a single-step extension of $\trace$
and $\resolve{\vhole{\thole}}{\vsu} \neq \resolve{\vhole{\thole}}{\vsu'}$
then $\vsu' = \vsu[\ehole_1 \mapsto v_1] \ldots [\ehole_n \mapsto v_n]$.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  $\vhole{\thole}$ does not change, so if the resolved values differ then
  $\vsu \neq \vsu'$.
  %
  Note that only \forcesym can update \vsu\
  and only the first case,
  when we extend \vsu\ with $[\ehole \mapsto v]$,
  does.
\end{proof}

\paragraph{Incompatible Types Are Wrong}
\emph{For all} types that are \emph{incompatible} with the
partial input type up to trace $\trace$, there exists a value
that will cause $f$ to get stuck in \emph{at most} $k$ steps,
where $k$ is the length of $\trace$.

\begin{lem}
\label{lem:k-stuck}
For all $t$,
if $\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu}$ and
   $\tincompat{t}{\ptype{\trace}{f}}$,
   then there exists a $v$ such that $\hastype{v}{t}$ and
   $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$ in at most
   $k$ steps, where $k$ is the length of $\trace$.
\end{lem}
\begin{proof}
  We can construct $v$ from $\trace$ by induction on $k$, the length of $\trace$.
  %
  The goal is to find the step at which the refinement of $\vhole{\thole}$,
  $\resolve{\vhole{\thole}}{\vsu}$ becomes incompatible with $t$.
  %
  The base case, $k = 0$ is trivial, $\vsu_0 = \emptysu$ and
  $\resolve{\vhole{\thole}}{\emptysu} = \vhole{\thole}$ which is compatible with
  \emph{any} type $t$.
  %
  In the inductive case, let
  $$\trace_{k-1} = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e_{k-1}}{\vsu_{k-1}}{\tsu_{k-1}}$$
  be the prefix of $\trace$ of length $k-1$.
  %
  Let us split cases on whether $\resolve{\vhole{\thole}}{\vsu_{k-1}}$ is compatible with $t$.
  %
  \begin{description}
  \item[Case $\vincompat{\resolve{\vhole{\thole}}{\vsu_{k-1}}}{t}$:]
    %
    The inductive hypothesis applies.
    %
  \item[Case $\vcompat{\resolve{\vhole{\thole}}{\vsu_{k-1}}}{t}$
        but  $\vincompat{\resolve{\vhole{\thole}}{\vsu_{k}}}{t}$:]
    %
    In this case we will show that $\resolve{\vhole{\thole}}{\vsu_{k-1}}$
    must contain some other hole $\ehole'$ which is instantiated with a
    concrete value at step $k$.
    %
    Furthermore, $\ehole'$ is instantiated in such a way that
    $\vincompat{\resolve{\vhole{\thole}}{\vsu_{k}}}{t}$.
    %
    Finally, we will show that if we had instantiated $\ehole'$ with a
    value such that
    $\vcompat{\resolve{\vhole{\thole}}{\vsu_{k}}}{t}$, the current
    step would have gotten $\stuck$.

    % By Lemma~\ref{lem:fixme} we know that
    % $\vcompat{\resolve{\vhole{\thole}}{\vsu}}{\resolve{\thole}{\tsu}}$.
    %
    By Lemma~\ref{lem:vsu-ext} we know that
    $\vsu_{k} = \vsu_{k-1}[\ehole_1 \mapsto v_1] \ldots [\ehole_n \mapsto v_n]$.
    %
    We will assume, without loss of generality, that
    $\vsu_{k} = \vsu_{k-1}[\ehole' \mapsto v']$.

    Next, we introduce a notion of \emph{value contexts}, defined
    analogously to evaluation contexts.
    %
    $$
      V ::= \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V}
    $$
    %
    Since $\vsu_{k-1}$ and $\vsu_{k}$ differ only in $\ehole'$ but the resolved
    values differ,
    $\resolve{\vhole{\thole}}{\vsu_{k-1}} = \vcompat{\invctx{\ehole'[\thole']}}{t}$
    and
    $\resolve{\vhole{\thole}}{\vsu_{k}} = \vincompat{\invctx{v'}}{t}$.
    %
    Now, consider any value $u$ such that $\vcompat{\invctx{u}}{t}$,
    we show by case analysis on the evaluation relation that
    $\invctx{u}$ could not make progress at this step, and thus
    $\steps{\eapp{f}{\invctx{u}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
  \end{description}
\end{proof}

% \paragraph{Incompatible Values Are Wrong}
%
% \emph{Any} value that is \emph{incompatible} with
% the partial input type upto trace $\trace$ will
% cause $f$ to get stuck in \emph{at most} $k$
% steps, where $k$ is the length of $\trace$.
% %
% \begin{lem}
% \label{lem:k-stuck}
%   For all $v$,
%   if \steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
%      \vincompat{v}{\ptype{\trace}{f}},
%   then
%      \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
%      in at most $k$ steps, where $k$ is the length of $\trace$.
% \end{lem}
% \begin{proof}
% By induction on $k$, the length of $\trace$.
% %
% Suppose {\vincompat{v}{\ptype{\trace}{f}}}.
% %
% We show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
% in at most $k$ steps.
% %
% The base case, $k = 0$ is trivial, $\trace$ is empty
% and so $\ptype{\trace}{f}$ is a hole that is compatible
% with \emph{every} value $v$.
% %
% In the inductive case, let
% $\trace' = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e'}{\vsu'}{\tsu'}$
% be the prefix of $\trace$ of length $k-1$.
% %
% Furthermore, let $s_{\trace} = \ptype{\trace}{f}$ and $s_{\trace'} = \ptype{\trace'}{f}$.
% %
% Let us split cases on whether $v$ is compatible with $s_{\trace'}$.
% %
% \begin{description}
% \item [Case \vincompat{v}{s_{\trace'}}:]
%   The inductive hypothesis applies.

% \item [Case $\vcompat{v}{s_{\trace'}}$ but $\vincompat{v}{s_{\trace}}$:]
%   Since $\vcompat{v}{s_{\trace'}}$ but $\vincompat{v}{s_{\trace}}$ we know
%   that $s_{\trace'} \neq s_{\trace}$.
%   By Lemma~\ref{lem:force-inst} we know that we must have
%   invoked \forcesym\ at step $k$.
%   %
%   The Preservation Lemma~\ref{lem:refine-partial} implies that
%   $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
%   specifically narrowed $s_{\trace'}$ to a type incompatible with $v$.
%   %
%   A case analysis of the evaluation rules shows that such an
%   invocation of \forcesym\ at step $k$
%   cannot succeed, \ie\ yields \stuck.
%    \RJ{some intuition needed --- seems like key step.}
% \end{description}
% \end{proof}

\begin{proof}[\textbf{Proof of Theorem~\ref{thm:soundness}}]
%
Suppose $\trace$ witnesses that $f$ gets stuck,
and let $t = \ptype{\trace}{f}$.
We show that \emph{all} types $s$ have stuck-inducing
values by splitting cases on whether the type is
compatible with $t$. %the partial type upto $\trace$.
%
\begin{description}
\item [Case \tcompat{s}{t}:]
  Let $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{\stuck}{\vsu}{\tsu}$.
  %
  The value $v = \resolve{\ehole}{\vsu}$ demonstrates that
  $\eapp{f}{v}$ gets stuck.
\item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, we can derive
  a $v$ from $\trace$ such that \hastype{v}{t} and $\eapp{f}{v}$ gets
  stuck.
% \item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$
%   such that \hastype{v}{t} demonstrates that $\eapp{f}{v}$ gets stuck.
%   % \ES{do we need to say anythign else?}
\end{description}
\end{proof}
